\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move

### 2025-06-11 writing/manuscript.Rmd
addressed KH's comments (need to make some figs)

```{r abstract, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have quantitatively estimated the effects of temperature on mammalian movement behaviour, but few have estimated the effects of future climate change. Consequently, it is still unclear how and where mammals will move in the 21^st^ century. We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada, between 1998 and 2023. We show that temperature is a strong determinant of when, how, and where mammals move, and we predict that climate change will impact mammals' movement rates and habitat selection as temperatures warm throughout the 21^st^ century. We found the effects of temperature on movement rates varied across species, so we suggest that species' estimated responses not be interpreted individually when making inferences about how climate change will impact ecological communities. In contrast, habitat selection strength decreased for all species within their observed ranges, suggesting that all six species may shift their range in the coming decades. As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effective and proactive conservation. We thus conclude by providing suggestions for conservation and future research while calling for a more holistic and collaborative framework."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

\noindent \textbf{References}: 141 (updated on 2025-05-16) <!-- counted by using movement ecology csl -->

@@

Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to respond to change in a proactive manner [@sawyer_identifying_2009; @mccain_assessing_2019]. The Canadian province of British Columbia is currently experiencing rapid widespread warming [@dierauer_climate_2021; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023]. As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness. Predicting how climate change will affect how, when, and where animals will move is necessary for identifying landscape dynamics and value will evolve throughout the century. Failing to understand how changes in temperature will affect mammalian movement behaviour and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.

@@

```{r dags,  message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graphs (DAGs) assumed for inferring the causal effects of temperature (Temp) on movement behavior. (A) DAG for the probability of moving, speed when moving, or distance traveled ($Y$) while accounting for the effects of time of day (TOD), day of year (DOY), and their interaction effects. Temperature directly affects $Y$, but the effects of temperature depend on the time of day and season. Time of day and day of year also affect $Y$ directly, but the effect of time of day changes throughout the year due to changes in day length and seasonality. (B) DAG for Relative Selection Strength (RSS) for percent forest cover (F), elevation (E), and distance from water (W). The RSS for of a given habitat depends on all three resources, and the selection for each resource is independent of the other two resources and dependent on temperature.", fig.height=4, fig.width=9, out.width='\\linewidth'}
library('dplyr')   # for data wrangling
library('ggplot2') # for fancy plots
library('ggdag')   # for directed acyclical graphs
library('cowplot') # for plots in grids

@@

```{r temp-qs, eval=FALSE}
library('dplyr')
d <- readRDS('data/hgam-speed-data.rds')
get_perc <- function(bools) {
  out <- as.character(round(mean(bools) * 100, 1))
  if(grepl('\\.', out)) {
    out <- paste0(out, ' %')
  } else {
    out <- paste0(out, '.0 %')
  }
  return(out)
}

bind_rows(
  d %>%
    rename(Species = species) %>%
    group_by(Species) %>%
    summarize(`$T < -20^\\\\circ C$` = get_perc(temp_c < -20),
              `$T > +20^\\\\circ C$` = get_perc(temp_c > +20)),
  d %>%
    summarize(Species = 'Total',
              `$T < -20^\\\\circ C$` = get_perc(temp_c < -20),
              `$T > +20^\\\\circ C$` = get_perc(temp_c > +20))) %>%
  knitr::kable(format = 'markdown', caption = paste0("Percentage of observed GPS locations (after data cleaning) with temperatures below -20\U00B0", "C or above 20\U00B0", "C."))
```

@@

\noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; Figs. B4-B6). Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models had good in-sample prediction (Fig. B7) and explained reasonably high proportions of the deviance (79.3% for the Gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses: see ch. 11 of @mcelreath_statistical_2020]. All species adapted their daily and seasonal movement behaviour to changes in temperature (Fig. \@ref(fig:dist)). The clearest example of this were cougars. In mid summer, they moved mostly in the evening if temperatures were cool and mostly in the early morning if temperatures were hot.   Throughout the year, they tended to move more when it was colder, but what counted as "colder" depended on the time of year. However, uncertainties around the estimated effects were often higher at extreme temperatures due to data scarcity. All species' estimated probability of moving decreased with sampling intervals approximately above 1 hour, and most species' estimated speed when moving decreased monotonically with sampling interval, although the estimated trends were highly uncertain for some species (Fig. B8).

@@

Species' relative selection strength (RSS) was generally strongest for elevation and weakest for forest cover, but species' RSS for resources depended on temperature. Changes in RSS with temperature were also strongest for elevation and generally weakest distance from water, but there were no common trends across all species for any of the three resources. All species, with the exception of cougars, exhibited a clear temperature-dependent shift in their preference for forest cover. At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou shifted towards more intermediate forest cover without much of a change in preference width. All species shifted elevationally with temperature, although boreal caribou did not exhibit as strong of a change. As temperatures rose, elk, mountain goats, and cougars increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation. Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources. Again, estimated RSS values were generally most uncertain at extreme temperatures (Fig. B14).

```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature. The grey contours indicate the extent of each species' observed locations. RSS values were re-centered and capped to $2^{\\pm 2}$ to improve readability."}
knitr::include_graphics('../figures/hrsf-surface-plots.png')
```

@@

\noindent Species varied in both the direction and magnitude of predicted change in distance traveled, but worse climate-change scenarios always corresponded to greater absolute changes in behaviour (Figs. \@ref(fig:t-dist), B9, and B10). Under the best-case scenario, absolute changes by 2100 were small (approximately 0% to 4%), while under the worst-case scenario absolute changes ranged from ~2% (grizzly bear) to ~25% (cougars), although the models did not explicitly account for changes in hibernation phenology. Throughout BC (Fig. \@ref(fig:s-dist)), all species showed little to no absolute change under the best-case scenario (approximately 0-4% relative to 2025, on average), and approximately 2-25% average absolute change under the worst-case scenario, but the predicted changes varied spatially due to heterogeneity in climate change. Again, absolute changes were smallest under the best-case scenario and greatest under the worst-case scenario for all species.

@@

```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent. Shaded areas indicate the 90% prediction interval within the range. Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs). The projections only account for changes movement in behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
```

@@

\noindent We have demonstrated that temperature is an important determinant of whether, how, and where mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources). Disentangling temperature from other drivers is important for predicting how changes in climate will affect mammalian movement when other drivers remain approximately constant. However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are exposed to increasingly frequent conditions that are both extreme and novel. Predicting mammals' responses at the data-scarce fringes of the surface plots in Figs. \@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear. At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures. As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023; @logares_black_2012; @anderson_black-swan_2017]. For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [@mortlock_sleep_2024] and likelihood to enter torpor [@fjelldal_nightly_2021]. Such changes may alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024]. At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021]. These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011].

Our ability to respond to current and future changes in climate is contingent on our ability to prepare for and predict change. However, predicting animal behaviour becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear and data are sparse. Consequently, we do not present our results as a definitive guide to how mammals in BC will respond to climate change. Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behaviour due to climate change. Additionally, communicating uncertainty in one's estimates is crucial in assessing risk probabilistically [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014]. Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [more commonly: "AI", see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024]. Mitigating climate change and its consequences will require extensive and widespread collaborative effort [@huang_achieving_2021]. Similarly, protecting a third of the world's ecosystems [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] appears less tangible as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023]. Achieving the "30 by 30" goal will require active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023]. Understanding the consequences of climate change on mammals' movement behaviour and spatial distribution is a first step towards anticipating and proactively responding to future changes in human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018]. In the following sections, we discuss the implications of our results in more detail. We then expand on consequences for conservation during the 21^st^ century and considerations for future studies.

@@

\noindent The lack of common effects of temperature across all species' movement rates indicates that preparing for future changes will require a variety of physiological and behavioral adaptations as ecological communities respond in complex and interconnected ways. Although our models do not account for explicit physiological or phenological changes, the tensor product terms in Fig. \@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns (most visible in cougars and grizzly bears). For example, when temperatures were above 0$^\circ$C, cougars moved most at night, but when temperatures were below 0$^\circ$C they tended to move more throughout the day. Throughout the year, they adapted their tolerance to temperature and moved less when it was relatively hot (for a given time of year), especially in spring and summer. The strong reduction in the mid-day movement rates of wolves, elk, cougars, and boreal caribou when summer temperatures were above 20$^\circ$C suggests that the increasingly common and intense heat waves across British Columbia will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020]. More work is necessary on quantifying interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024]. The increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022]. However, not all species may be able to adapt at the current rate of climate change [@williams_range_2018; @hetem_responses_2014], especially since landscape changes will likely depend not only on the direct effect of temperature but also on cascading changes in the availability and unpredictability of resources [@mclellan_effect_2015; @pigeon_drivers_2016; @mezzini_how_2025].

@@

As with the effects of temperature on movement rates, the lack of common effects of temperature across all species indicates that efficient and proactive habitat conservation will require taxonomic nuance. In particular, our HRSFs did not account for any explicit physiological responses to temperature that may impact movement, such as changes in the phenology of plants, ice, or mammalian physiology or behaviour (e.g., moulting, migration), other than as direct responses to temperature. Consequently, the behavioural changes we present should be interpreted carefully. While some seasonal cycles do not depend directly on temperature and may thus not be impacted directly by climate change [such as those driven by photoperiod, see @liu_effects_2022], the predictions we present also rely on the assumption that mammals will not adapt these seasonal cycles to temperature-dependent cues, such as weather [@xu_frost_2019] or forage availability [@middleton_green-wave_2018], which may not be correct. Additionally, it is important to stress the distinction between habitat selection and quality. While animals tend to select for high-quality habitats [@kacelnik_ideal_1992; @wright_survival_2024], high selection strength can also be a consequence of an ecological trap [@zuniga-palacios_what_2021; @swearer_overview_2021].

@@

Achieving the goals laid out by the "30 by 30" conservation initiative [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action. In the case of Canada, only 13.7% of its terrestrial areas and inland waters were protected at the end of 2023 [@environment_and_climate_change_canada_canadian_2024; @minister_of_environment_and_climate_change_act_2024]. Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [e.g., @falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] while also actively preparing for future change [@desjardins_looking_2020; @hessami_indigenizing_2021]. The large range in predicted changes by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig. \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates. The width of the prediction intervals in Fig. \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas. Accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019]. This complexity results in the great variation among the responses of different species (and individuals), especially as populations a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022]. As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].

@@

>>>>

```{r abstract, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have quantitatively estimated the effects of temperature on mammalian movement behaviour, but few have estimated the effects of future climate change. Consequently, it is still unclear how mammals will adapt their movement behavior in the 21^st^ century. We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada, between 1998 and 2023. We show that temperature is a strong determinant of when, how, and where mammals move, and we predict that climate change will impact mammals' movement rates and habitat selection as temperatures warm throughout the 21^st^ century. We found the effects of temperature on movement rates varied across species, so we suggest that species' estimated responses be interpreted holistically when making inferences about how climate change will impact ecological communities. In contrast, habitat selection strength decreased for all species within their observed ranges, suggesting that all six species may shift their range in the coming decades. As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effective and proactive conservation. We thus conclude by providing suggestions for conservation and future research, with particular attention to the study design, statistical modeling, and forecasting. We provide detailed `R` code and appendices so that other researchers can implement our methods on their own datasets."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

\noindent \textbf{References}: 142 (updated on 2025-06-11) <!-- counted by using movement ecology csl -->

@@

Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to respond to change in a proactive manner [@sawyer_identifying_2009; @mccain_assessing_2019]. The Canadian province of British Columbia is currently experiencing rapid widespread warming [@dierauer_climate_2021; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023]. As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness. Predicting how climate change will affect how, when, and where animals will move is necessary for identifying landscape dynamics and value will evolve throughout the century. Failing to understand how changes in temperature will affect mammalian movement behaviour and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.

@@

```{r dags,  message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graphs (DAGs) assumed for inferring the causal effects of temperature (Temp) on movement behavior. (A) DAG for the probability of moving, speed when moving, or distance traveled ($Y$) while accounting for the effects of time of day (TOD), day of year (DOY), and their interaction effects. Temperature directly affects $Y$, but the effects of temperature depend on the time of day and season. Time of day and day of year also affect $Y$ directly, but the effect of time of day changes throughout the year due to changes in day length and seasonality. (B) DAG for Relative Selection Strength (RSS) for percent forest cover (F), elevation (E), and distance from water (W). The RSS for of a given habitat depends on all three resources, and the selection for each resource is independent of the other two resources and dependent on temperature.", fig.height=4, fig.width=10, out.width='\\linewidth'}
library('dplyr')   # for data wrangling
library('ggplot2') # for fancy plots
library('ggdag')   # for directed acyclical graphs
library('cowplot') # for plots in grids

@@

```{r temp-qs}
library('dplyr')
d <- readRDS('../data/hgam-speed-data.rds')
get_perc <- function(bools) {
  out <- as.character(round(mean(bools) * 100, 1))
  if(grepl('\\.', out)) {
    out <- paste0(out, ' %')
  } else {
    out <- paste0(out, '.0 %')
  }
  return(out)
}

bind_rows(
  d %>%
    rename(Species = species) %>%
    group_by(Species) %>%
    summarize(`T < -20\\textdegree{} C` = get_perc(temp_c < -20),
              `T > +20\\textdegree{} C` = get_perc(temp_c > +20)),
  d %>%
    summarize(Species = 'Total',
              `T < -20\\textdegree{} C` = get_perc(temp_c < -20),
              `T > +20\\textdegree{} C` = get_perc(temp_c > +20))) %>%
  knitr::kable(format = 'markdown', caption = paste0("Percentage of observed GPS locations (after data cleaning) with temperatures below -20\U00B0", "C or above 20\U00B0", "C."))
```

@@

\noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; Figs. B4-B6). Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models had good in-sample prediction (Fig. B7) and explained reasonably high proportions of the deviance (79.3% for the Gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses: see ch. 11 of @mcelreath_statistical_2020]. All species adapted their daily and seasonal movement behaviour to changes in temperature (Fig. \@ref(fig:dist)). The clearest example of this was cougars. In mid summer, they moved mostly in the evening if temperatures were cool and mostly in the early morning if temperatures were hot. Throughout the year, they tended to move more when it was colder, but what they perceived as "colder" depended on the time of year. However, uncertainties around the estimated effects were often higher at extreme temperatures due to data scarcity. All species' estimated probability of moving decreased with sampling intervals approximately above 1 hour, and most species' estimated speed when moving decreased monotonically with sampling interval, although the estimated trends were highly uncertain for some species (Fig. B8).

@@

Species' relative selection strength (RSS) was generally strongest for elevation and weakest for forest cover, but species' RSS for resources depended on temperature (Fig. \@ref(fig:hrsfs)). Changes in RSS with temperature were also strongest for elevation and generally weakest distance from water, but there were no common trends across all species for any of the three resources. All species, with the exception of cougars, exhibited a clear temperature-dependent shift in their preference for forest cover. At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou shifted towards more intermediate forest cover without much of a change in preference width. All species shifted elevationally with temperature, although boreal caribou's selection strength for elevation was less sensitive to elevation. Overall, as temperatures rose, elk, mountain goats, and cougars increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation. Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources. Again, estimated RSS values were generally most uncertain at extreme temperatures (Fig. B14).

```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature. The grey contours indicate the extent of each species' observed data. RSS values were re-centered and capped to $2^{\\pm 2}$ to improve readability."}
knitr::include_graphics('../figures/hrsf-surface-plots.png')
```

@@

\noindent Species varied in both the direction and magnitude of predicted change in distance traveled, but worse climate-change scenarios always corresponded to greater absolute changes in behaviour (Figs. \@ref(fig:t-dist), B9, and B10). Under the best-case scenario, absolute changes by 2100 were small (approximately 0% to 4%), while under the worst-case scenario absolute changes ranged from ~2% (grizzly bear) to ~25% (cougars). However, while the models estimated that grizzly bears would move substantially less (if at all) in winter (Figs. \@ref(fig:dist), B4-B6), the models did not explicitly account for changes in hibernation phenology. Consequently, the climate change projections do not explicitly account for changes in energetic needs or physiology. Throughout BC (Fig. \@ref(fig:s-dist)), all species showed little to no absolute change under the best-case scenario (approximately 0-4% relative to 2025, on average), and approximately 2-25% average absolute change under the worst-case scenario, but the predicted changes varied spatially due to heterogeneity in climate change. Again, absolute changes were smallest under the best-case scenario and greatest under the worst-case scenario for all species.

@@

```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent. Shaded areas indicate the 90% prediction interval within the range. Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs). The projections only account for changes in movement in behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
```

@@

\noindent We have demonstrated that temperature is an important determinant of whether, how, and where large mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources). Disentangling temperature from other drivers is important for predicting how changes in climate will affect mammalian movement when other drivers remain approximately constant. However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are exposed to increasingly frequent conditions that are both extreme and novel.

Predicting mammals' responses at the data-scarce fringes of the surface plots in Figs. \@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear. At warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures. As extreme temperature events become more frequent and prolonged throughout the year [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023; @logares_black_2012; @anderson_black-swan_2017]. For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [@mortlock_sleep_2024] and likelihood to enter torpor [@fjelldal_nightly_2021].

Changes in animal movement behavior may also alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024]. At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being preyed on [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021]. These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011].

Our ability to respond changes in climate is contingent on our ability to prepare for and predict change. However, predicting animal behaviour becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear and data are sparse. Consequently, we do not present our results as a definitive guide to how mammals in BC will respond to climate change. Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behaviour due to climate change. Additionally, communicating uncertainty in one's estimates is crucial in assessing risk probabilistically, and appropriate measures of uncertainty require careful and robust modeling [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014].

Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [more commonly: "AI", see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024]. Mitigating climate change and its consequences will require extensive and widespread collaborative effort [@huang_achieving_2021]. Similarly, protecting a third of the world's ecosystems [@convention_on_biological_diversity_154_2022] appears less achievable as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023]. Achieving the "30 by 30" goal will require international collaboration as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023]. Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018]. In the following sections, we discuss the implications of our results in more detail. We then expand on consequences for conservation during the 21^st^ century and considerations for future studies.

@@

\noindent The heterogeneity mammals' responses to temperature suggests that ecological communities will respond to change in complex and interconnected ways. Although our models do not account for explicit physiological or phenological changes, the surfaces in Fig. \@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns (most visible in cougars and grizzly bears). For example, when temperatures were above 0$^\circ$C, cougars moved most at night, but when temperatures were below 0$^\circ$C they tended to move more throughout the day. Throughout the year, they adapted their tolerance to temperature and moved less when it was relatively hot (for a given time of year), especially in spring and summer. The strong reduction in the mid-day movement rates of wolves, elk, cougars, and boreal caribou when summer temperatures were above 20$^\circ$C suggests that the increasingly common and intense heat waves across British Columbia will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020]. More work is necessary on quantifying interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024]. The increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022]. However, not all species may be able to adapt at the current rate of climate change [@williams_range_2018; @hetem_responses_2014], especially since landscape changes will likely depend not only on the direct effect of temperature but also on cascading changes in the availability and unpredictability of resources [@mclellan_effect_2015; @pigeon_drivers_2016; @mezzini_how_2025].

@@

As with the effects of temperature on movement rates, the hererogeneity of temperature effects on species' habitat selection indicates that efficient and proactive habitat conservation will require taxonomic nuance. In particular, our HRSFs did not account for any explicit physiological responses to temperature that may impact movement, such as changes in the phenology of plants, fire, ice, or mammalian physiology (e.g., moulting, fat storages) or behaviour (e.g., migration, food caching), other than as direct responses to temperature. Consequently, the behavioural changes we present should be interpreted carefully. While some seasonal cycles do not depend directly on temperature and may thus not be impacted directly by climate change [such as those driven by photoperiod, see @liu_effects_2022], the predictions we present also rely on the assumption that mammals will not adapt these seasonal cycles to temperature-dependent cues, such as weather [@xu_frost_2019] or forage availability [@middleton_green-wave_2018], which may not be correct. Additionally, it is important to stress the distinction between habitat selection and quality. While animals tend to select for high-quality habitats [@kacelnik_ideal_1992; @wright_survival_2024], high selection strength can also be a consequence of an ecological trap [@zuniga-palacios_what_2021; @swearer_overview_2021].

@@

Achieving the goals laid out by the "30 by 30" conservation initiative [@convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action. In the case of Canada, only 13.7% of its terrestrial areas and inland waters were protected at the end of 2023 [@environment_and_climate_change_canada_canadian_2024; @minister_of_environment_and_climate_change_act_2024]. Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [@falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] while also actively preparing for future change [@desjardins_looking_2020; @hessami_indigenizing_2021].

The diversity in predicted changes in movement behavior by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig. \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates. For example, the width of the prediction intervals in Fig. \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas. Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024]. Additionally, accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019]. This complexity results in the great variation among the responses of different species (and individuals), especially as populations undergo a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022]. As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].

@@

Increasing temperatures increased RSS for both elevations near ~500 m as well as > 1,000 m, but such a selection is only possible if: (1) such habitats exist in the animals' range, and (2) crossing the 500 m to > 1,000 m regions is safe and attractive
